---
title: "KPMP TAL Data"
subtitle: "Exploration of KRT19 Expression"
author: "Jonathan Nelson"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document:
    toc: yes
    toc_depth: 4
    toc_float: 
      collapsed: false
      smooth_scroll: false
    number_sections: yes
    theme: bootstrap
    df_print: paged
    code_folding: hide
    highlight: pygments
editor_options: 
  chunk_output_type: inline
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r packages, echo=T, error=FALSE, message=FALSE, warning=FALSE}

if (!require("dplyr")) {install.packages("dplyr"); require("dplyr")}
if (!require("Seurat")) {install.packages("Seurat"); require("Seurat")}
if (!require("SeuratDisk")) {install.packages("SeuratDisk"); require("SeuratDisk")}
if (!require("knitr")) {install.packages("knitr"); require("knitr")}
if (!require("ggplot2")) {install.packages("ggplot2"); require("ggplot2")}
if (!requireNamespace('BiocManager', quietly = TRUE)) {install.packages('BiocManager'); require("BiocManager")}
if (!require("here")) {install.packages("here"); require("here")}
if (!require("paletteer")) {install.packages("paletteer"); require("paletteer")} # color palette
if (!require("grDevices")) {install.packages("grDevices"); require("grDevices")} # for grDevices palette
if (!require("tidyverse")) {install.packages("tidyverse"); require("tidyverse")} # for data frame transformation
if (!require("tibble")) {install.packages("tibble"); require("tibble")} # for table transformation
if (!require("geneName")) {install.packages("geneName"); require("geneName")}
if (!require("ggrepel")) {install.packages("ggrepel"); require("ggrepel")}
if (!require("gghighlight")) {install.packages("gghighlight"); require("gghighlight")}
if (!require("ggpmisc")) {install.packages("ggpmisc"); require("ggpmisc")}
if (!require("ggupset")) {install.packages("ggupset"); require("ggupset")}
if (!require("RColorBrewer")) {install.packages("RColorBrewer"); require("RColorBrewer")}
if (!require("viridis")) {install.packages("viridis"); require("viridis")}

library(SeuratData)
library(openxlsx)
library(gplots)
library(ggvenn)

```


# Introduction

This file was made to explore the expression of KRT19 in TAL snRNASeq data from the KPMP 2021 snRNAseq dataset for Vidhi Dalal.

# 2021 KPMP Object

## Load KPMP object (2021 data file)

This is the KPMP object that was originally formatted by Xiao-Tong Su (manually annotated meta-data) and then subsetted by Jessica Bahena Lopez for her TAL heterogenity manuscript. 

```{r, echo=TRUE, warning=FALSE, error=FALSE, message=FALSE, fig.align = 'center'}

KPMP.TAL <- readRDS(here("KPMP.TAL_2.rds")) 

head(KPMP.TAL@meta.data)

KPMP.TAL@meta.data$subclass.l2 <- factor(KPMP.TAL@meta.data$subclass.l2, levels = c("C-TAL", "M-TAL", "aTAL1", "aTAL2", "dC-TAL", "dM-TAL", "MD"))


DimPlot(KPMP.TAL, group.by = "subclass.l1")

DimPlot(KPMP.TAL, group.by = "subclass.l2", label = T)

KPMP.TAL

```

## VlnPlot -> Heatmap

```{r, echo=TRUE, warning=FALSE, error=FALSE, message=FALSE, fig.align = 'center', fig.height=10}

gene_list <- list(c("SLC12A1", "CRYAB", "KRT19", "MMP7", "CLU", "SPP1"))

gene_unlist <- unlist(c("SLC12A1", "CRYAB", "KRT19", "MMP7", "CLU", "SPP1"))

VlnPlot(KPMP.TAL, features = gene_unlist, group.by = "subclass.l2", pt.size = 0.1)

```

```{r, echo=TRUE, warning=FALSE, error=FALSE, message=FALSE, fig.align = 'center', fig.width=10}

# Create a dataframe with the average expression for the list of genes

df <- AverageExpression(object = KPMP.TAL, features = gene_unlist, group.by = 'subclass.l2')$RNA

df

# Calculate Z score for each row (could skip if you want raw expression)

df <- t(scale(t(df)))

# convert df to tidy format

df_tidy <- df %>%
  as.data.frame() %>%
  rownames_to_column(var = "Gene") %>%
  pivot_longer(cols = -Gene, names_to = "subclass.l2", values_to = "Expression")
               
# Graph with geom_tile

f1 <- ggplot(df_tidy, aes(x = subclass.l2, y = Gene, fill = Expression)) +
  geom_tile()

f1

# Make pretty with ggplot2

f2 <- ggplot(df_tidy, aes(x = subclass.l2, y = Gene, fill = Expression)) +
  geom_tile() +
  scale_fill_distiller(palette = "RdYlBu") +
  theme_minimal()

f2

# Reorder the genes by factoring

df_tidy$Gene <- factor(df_tidy$Gene, levels = c("CRYAB",  "MMP7", "CLU", "SPP1", "KRT19","SLC12A1"))

# Reorder sublcass.l2 by factoring

df_tidy$subclass.l2 <- factor(df_tidy$subclass.l2, levels = c("C-TAL", "M-TAL", "aTAL1", "aTAL2", "dC-TAL", "dM-TAL", "MD"))

```



## DimPlots of TAL by groups

```{r, echo=TRUE, warning=FALSE, error=FALSE, message=FALSE, fig.align = 'center', fig.width=10}


KPMP.TAL@meta.data$Enrollment.Category <- factor(KPMP.TAL@meta.data$Enrollment.Category, levels = c("Healthy Reference", "AKI", "CKD"))

DimPlot(KPMP.TAL, group.by = "subclass.l2", label = T, split.by = "Enrollment.Category")

KPMP.TAL@meta.data


```

```{r, echo=TRUE, warning=FALSE, error=FALSE, message=FALSE, fig.align = 'center', fig.width=15}


KPMP.TAL@meta.data$Proteinuria..mg...Binned. <- factor(KPMP.TAL@meta.data$Proteinuria..mg...Binned., levels = c("<150 mg/g cr", "150 to <500 mg/g cr", "500 to <1000 mg/g cr", ">=1000 mg/g cr", ""))

DimPlot(KPMP.TAL, group.by = "subclass.l2", label = T, split.by = "Proteinuria..mg...Binned.")

```


## FeaturePlots of TAL for KRT19

```{r, echo=TRUE, warning=FALSE, error=FALSE, message=FALSE, fig.align = 'center'}

FeaturePlot(KPMP.TAL, "KRT19", order = T)

```


```{r, echo=TRUE, warning=FALSE, error=FALSE, message=FALSE, fig.align = 'center', fig.width=10}

FeaturePlot(KPMP.TAL, "KRT19", split.by = "Enrollment.Category", order = T)

```

```{r, echo=TRUE, warning=FALSE, error=FALSE, message=FALSE, fig.align = 'center', fig.width=15}


FeaturePlot(KPMP.TAL, "KRT19", split.by = "Proteinuria..mg...Binned.", order = T)

```


## VlnPlots of TAL for KRT19

```{r, echo=TRUE, warning=FALSE, error=FALSE, message=FALSE, fig.align = 'center'}

VlnPlot(KPMP.TAL, "KRT19", group.by = "Enrollment.Category")

VlnPlot(KPMP.TAL, "KRT19", group.by = "Proteinuria..mg...Binned.")

VlnPlot(KPMP.TAL, "KRT19", group.by = "subclass.l2")

```

## DotPlots of TAL for KRT19

### Scaled

```{r, echo=TRUE, warning=FALSE, error=FALSE, message=FALSE, fig.align = 'center', fig.width=6}

DotPlot(
  KPMP.TAL,
  assay = NULL,
  features = "KRT19",
  cols = c("lightgrey", "blue"),
  col.min = -2.5,
  col.max = 2.5,
  dot.min = 0,
  dot.scale = 6,
  idents = NULL,
  group.by = "Enrollment.Category",
  split.by = NULL,
  cluster.idents = FALSE,
  scale = T,
  scale.by = "radius",
  scale.min = NA,
  scale.max = NA
)


DotPlot(
  KPMP.TAL,
  assay = NULL,
  features = "KRT19",
  cols = c("lightgrey", "blue"),
  col.min = -2.5,
  col.max = 2.5,
  dot.min = 0,
  dot.scale = 6,
  idents = NULL,
  group.by = "subclass.l2",
  split.by = NULL,
  cluster.idents = FALSE,
  scale = T,
  scale.by = "radius",
  scale.min = NA,
  scale.max = NA
)

```


### Unscaled

```{r, echo=TRUE, warning=FALSE, error=FALSE, message=FALSE, fig.align = 'center', fig.width=6}

DotPlot(
  KPMP.TAL,
  assay = NULL,
  features = "KRT19",
  cols = c("lightgrey", "blue"),
  col.min = -2.5,
  col.max = 2.5,
  dot.min = 0,
  dot.scale = 6,
  idents = NULL,
  group.by = "Enrollment.Category",
  split.by = NULL,
  cluster.idents = FALSE,
  scale = F,
  scale.by = "radius",
  scale.min = NA,
  scale.max = NA
)

DotPlot(
  KPMP.TAL,
  assay = NULL,
  features = "KRT19",
  cols = c("lightgrey", "blue"),
  col.min = -2.5,
  col.max = 2.5,
  dot.min = 0,
  dot.scale = 6,
  idents = NULL,
  group.by = "subclass.l2",
  split.by = NULL,
  cluster.idents = FALSE,
  scale = F,
  scale.by = "radius",
  scale.min = NA,
  scale.max = NA
)

```


# Correlating KRT19 cells with VD New TAL

## Add meta.data column called "KRT19status" depending on whether a cell expressed KRT19 or not

```{r echo=TRUE, warning=FALSE, error=FALSE, message=FALSE, fig.align = 'center'}

KPMP.TAL@meta.data$KRT19status <- ifelse(GetAssayData(KPMP.TAL, assay = "RNA", slot = "data")["KRT19", ] > 0, "yes", "no")

DimPlot(KPMP.TAL, group.by = "KRT19status", label = T)

VlnPlot(KPMP.TAL, "KRT19", assay = "RNA", group.by = "KRT19status")

Idents(KPMP.TAL) <- "KRT19status"

DimPlot(KPMP.TAL)

KRT19.DEGs <- FindMarkers(KPMP.TAL, ident.1 = "yes", ident.2 = "no")

KRT19.DEGs

```
```{r, echo=TRUE, warning=FALSE, error=FALSE, message=FALSE, fig.align = 'center', fig.width=10}

VD.DEGs <- read.xlsx(here("DEG_NewTAL_vs_TAL.xlsx"))

# change x1 to "gene"

VD.DEGs <- VD.DEGs %>% rename(gene = X1)

x.markers <- VD.DEGs


x.markers_tb <- x.markers %>% data.frame() %>% filter(p_val_adj < 0.05) %>% filter(abs(avg_log2FC) > .5) %>%  as_tibble()

top6 <- list(head(x.markers_tb$gene))

x.markers_tb_H <- mousegnameConverter(x.markers_tb, "gene")

x.markers_tb_H = x.markers_tb_H[order(x.markers_tb_H[,"avg_log2FC"], decreasing = TRUE),]
x.markers_tb_H

# move the KRT19.DEG rownames to be a column called "gene"

y.markers <- KRT19.DEGs

y.markers_tb <- y.markers %>%
  data.frame() %>%
  rownames_to_column(var="gene") %>%
  filter(p_val_adj < 0.05) %>%
  filter(abs(avg_log2FC) > .5) %>% 
  as_tibble()

    #X-Y DEGs Intersection Table
    
    xy.comp.TAL <- inner_join(x.markers_tb_H, y.markers_tb, by = "gene")
    
    xy.comp.TAL 
    
    #Set Range for Far Right Data Points
    df.upper <- subset(xy.comp.TAL, avg_log2FC.x > -.32 & avg_log2FC.y > -.32)
    #Set Range for Far Left Data Points
    df.lower <- subset(xy.comp.TAL, avg_log2FC.x < 0.32 & avg_log2FC.y < .32)
    
    xy.comp.TAL.plot <- ggplot(xy.comp.TAL, aes(x = avg_log2FC.x, y = avg_log2FC.y, label=gene)) +
      theme_classic() +
      geom_point(
        color=dplyr::case_when(
          (xy.comp.TAL$avg_log2FC.x > 1 & xy.comp.TAL$avg_log2FC.y > 1) ~ "#1b9e77", #sets color for df.upper points
          (xy.comp.TAL$avg_log2FC.x < -1 & xy.comp.TAL$avg_log2FC.y < -1) ~ "#d95f02", #sets color for df.lower points
          TRUE ~ "black")) +
      geom_text_repel(data=rbind(df.upper, df.lower),
                      segment.sixy.comp.TALe  = 0.2, #<--!! what is this?? !!--
                      segment.color = "grey50") +
      geom_smooth (method=lm) +
      labs(
        title = paste("Correlation of Log2FC Values of DEGs from",
                      "Mouse New TAL", "and",
                      "KPMP KRT19+ Cells", sep = " "), 
        x = paste("Average log2FC", "Mouse"), 
        y = paste("Average log2FC", "KPMP")
      ) +
      stat_poly_eq(aes(label = paste(..rr.label.., sep = "~~~")),
                   label.x.npc = "left", label.y.npc = 0.90, #set the position of the eq
                   rr.digits = 3)
    
    print(xy.comp.TAL.plot)
    
    print(nrow(xy.comp.TAL))

```

## Corerelation Test

```{r, echo=TRUE, warning=FALSE, error=FALSE, message=FALSE, fig.align = 'center'}

cor_result <- cor.test(xy.comp.TAL$avg_log2FC.x, xy.comp.TAL$avg_log2FC.y, method = "pearson")
cor_result

```


```{r echo=TRUE, warning=FALSE, error=FALSE, message=FALSE, fig.align = 'center'}

Sys.time()

sessionInfo()

```
